European or American? A re-examination of the transatlantic influence over the IMF

### 2025-07-30 manuscript/manuscript.qmd
switch to Brirtish spelling

Compared to the extensive literature on U.S. dominance in the IMF, scholarship on European influence is relatively limited and often fragmented. Many studies underscore that EU member states are institutionally dispersed across multiple constituencies, hindering coherent collective action [@smaghi2004; @miescu2014; @thacker1999]. Still, other works argue that European countries wield significant structural advantages—including leadership appointments, voting power, and coordination mechanisms—that enable them to shape lending outcomes in line with their interests [@copelovitch2010a; @keating2024; @lutz2014; @stone2004].

European influence reflects a strategic combination of formal power and coordinated action. EU countries collectively control over 30% of IMF voting rights, and if they were to unify, they could potentially counterbalance U.S. veto power [@brandner2009a; @leech2005a; @smaghi2004]. This is reinforced by the long-standing transatlantic norm allocating IMF leadership to a European national [@kahler2001]. Through platforms such as EURIMF and SCIMF, EU members have harmonized policy positions and influenced loan design. During the Eurozone crisis, France and Germany pushed for IMF involvement to stabilize the region while embedding austerity-oriented conditionality that protected their banking systems [@hodson2014; @kickert2019a; @veron2016]. Lending terms were shaped to reflect creditor priorities, with evidence that conditionality was both intensified or relaxed based on geopolitical alignment [@lutz2014; @heimberger2025, p. 9; @dreher2015]. Beyond Europe, the UK and France have used IMF concessional lending in Africa to maintain post-colonial ties and commercial access [@stone2004].

Despite Europe’s institutional advantages, scholars remain skeptical of its actual effectiveness in the IMF. Internal divergences—such as France’s growth-oriented stance versus Germany’s fiscal conservatism—combined with fragmented constituencies, often impede coherent action and reduce policy influence [@miescu2014; @dreher2015; @copelovitch2010a]. Moreover, Europe’s disproportionate voting power relative to its population raises legitimacy concerns and is seen as disadvantaging developing countries [@woodward2024; @leech2005a]. While proposals to consolidate EU seats aim to improve efficiency and counterbalance U.S. dominance, critics warn such reforms may merely reinforce dominance by a few large European states rather than broaden representation [@leech2005a; @copelovitch2010a].

While several studies examine European involvement in IMF operations, most focus on individual cases or regional crises, lacking generalizable conclusions. Existing literature often treats Europe as a secondary actor or subsumes it under undifferentiated “Western” influence, thereby overlooking intra-Western variation. By analytically separating U.S. and European roles, our study addresses these gaps and offers a comparative framework for understanding divergent patterns of influence within global financial governance.

>>>>

Compared to the extensive literature on US dominance in the IMF, scholarship on European influence is relatively limited and often fragmented. Many studies underscore that EU member states are institutionally dispersed across multiple constituencies, hindering coherent collective action [@smaghi2004; @miescu2014; @thacker1999]. Still, other works argue that European countries wield significant structural advantages—including leadership appointments, voting power, and coordination mechanisms—that enable them to shape lending outcomes in line with their interests [@copelovitch2010a; @keating2024; @lutz2014; @stone2004].

European influence reflects a strategic combination of formal power and coordinated action. EU countries collectively control over 30% of IMF voting rights, and if they were to unify, they could potentially counterbalance US veto power [@brandner2009a; @leech2005a; @smaghi2004]. This is reinforced by the long-standing transatlantic norm allocating IMF leadership to a European national [@kahler2001]. Through platforms such as EURIMF and SCIMF, EU members have harmonised policy positions and influenced loan design. During the Eurozone crisis, France and Germany pushed for IMF involvement to stabilise the region while embedding austerity-oriented conditionality that protected their banking systems [@hodson2014; @kickert2019a; @veron2016]. Lending terms were shaped to reflect creditor priorities, with evidence that conditionality was both intensified or relaxed based on geopolitical alignment [@lutz2014; @heimberger2025, p. 9; @dreher2015]. Beyond Europe, the UK and France have used IMF concessional lending in Africa to maintain post-colonial ties and commercial access [@stone2004].

Despite Europe’s institutional advantages, scholars remain skeptical of its actual effectiveness in the IMF. Internal divergences—such as France’s growth-oriented stance versus Germany’s fiscal conservatism—combined with fragmented constituencies, often impede coherent action and reduce policy influence [@miescu2014; @dreher2015; @copelovitch2010a]. Moreover, Europe’s disproportionate voting power relative to its population raises legitimacy concerns and is seen as disadvantaging developing countries [@woodward2024; @leech2005a]. While proposals to consolidate EU seats aim to improve efficiency and counterbalance US dominance, critics warn such reforms may merely reinforce dominance by a few large European states rather than broaden representation [@leech2005a; @copelovitch2010a].

While several studies examine European involvement in IMF operations, most focus on individual cases or regional crises, lacking generalisable conclusions. Existing literature often treats Europe as a secondary actor or subsumes it under undifferentiated “Western” influence, thereby overlooking intra-Western variation. By analytically separating US and European roles, our study addresses these gaps and offers a comparative framework for understanding divergent patterns of influence within global financial governance.
